---
title: "Measures Against Covid-19 in Greece: A Small Analysis"
author: "Georgios Lamprou"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
bibliography: references.bib
---

```{r setup, include=FALSE}
# Set global options for all code chunks
knitr::opts_chunk$set(
  fig.width = 8,         # Default figure width in inches
  fig.height = 5.5,        # Default figure height in inches
  fig.align = "center"   # Center align all plots
)
```

```{=html}
<style>
  body {
    font-size: 16px; /* Increase the base font size */
    text-align: justify;
  }
</style>
```

### Introduction

Five years after the start of the Covid-19 pandemic, the world continues to live in a situation of symbiosis with the virus Sars-Cov-2. Even in our days, a lot of people suffer from coronavirus for first or second or third time, a phenomenon which called reinfection [@reinfect2024]. However, in the difficult triennial (2020 - 2023), governments around the world were constantly making decisions to introduce or revise a wide variety of measures in order to mitigate the pandemic and protect the human life [@güner2021].

The same approach was adopted in Greece with general effectiveness. Other nations admired the Greek effort and tactics to deal with the pandemic. In any circumstance, the immediate strict measures' application on a nationwide as well as on a local level and the discipline of people played a huge role for this kind of success [@papathanassopoulos2023].

Based on the above, this article aims to present a small analysis of the measures against Covid-19 in Greece relative to different sectors. It is necessary to refer that the database "Public Health and Social Measures in Response to COVID-19" from World Health Organization (WHO) is used for the acquisition of data. To explore this useful database for European Region, the corresponding link is provided here: <https://phsm.euro.who.int/covid-19>. Furthermore, it is worth to refer that the current article consists of 5 sections.

### Tools

The visualizations created with R Programming Language and the technology of Quarto Documents used to write the article.

### Measures During the Pandemic Years

To begin with, as seems in the graph below, the majority of the measures recorded on 2021 (393 records) and the minority of them on 2024 (3 records).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Libraries
  library(dplyr)
  library(readxl)
  library(ggplot2)
  library(ggthemes)
  library(plotly)
  library(collapsibleTree)
  library(paletteer)
  library(lisa)
  library(ggrepel)
  library(patchwork)
  library(GGally)

  data <- read_xlsx("phsm_database_04032024_greece.xlsx", col_names = TRUE)
  
  # Data Processing and Transformations
  
  data$`Start of measure` <- as.Date.POSIXct(data$`Start of measure`)
  data$`End of measure` <- as.Date.POSIXct(data$`End of measure`)
  data$`Date of access` <- as.Date.POSIXct(data$`Date of access`)
  
  
  # Different Statistics - Graphs
  
  # Number of Measures' Record during the years
  
  freq_of_measures_based_on_year_entry <- data %>%
    dplyr::select(`Entry year`)%>%
    count(`Entry year`)
  
  freq_of_measures_based_on_year_entry$`Entry year` <- as.factor(freq_of_measures_based_on_year_entry$`Entry year`)
  
  g0 <- ggplot(freq_of_measures_based_on_year_entry, aes(x= `Entry year`, y=n, fill=`Entry year`)) +
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Records of Measures by Year") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Year") +
    ylab("Number of Records") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
g0  
```

This point is not strange. Instead, is very logical for 2 main reasons:

1.  2021 was the next year of 2020, the pandemic's official starting year [@jebril2020] while an important amount of 2021's days included in the second lockdown which started in November 2020 and finished 210 later [@papazisis2021].

2.  In 2021 many measures which were related to vaccinations and Delta mutation of Sars-Cov-2 virus, were introduced for 1st time or modified, [@Liossi2023; @Tsiambas2021].

The above result could also be framed by the following time series graph. In this, it is observed that the most of the measures started in 2021.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
  data_for_timeline_of_measure_started <- data %>%
    dplyr::select(`Start of measure`) %>%
    count(`Start of measure`)
  
  data_for_timeline_of_measure_ended <- data %>%
    dplyr::select(`End of measure`) %>%
    count(`End of measure`)
  
  g5 <- ggplot(data_for_timeline_of_measure_started, aes(x=`Start of measure`, y=n))+
    geom_line(stat = 'identity',color='darkred') +
    ggtitle("Total Number of Measures Started by Day") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Date") +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12") 
   
  ggplotly(g5)
```

To continue, in the next bar plot, the fact that in the triennial 2020, 2021, 2022 most of the measures belong to the category "Social and physical distancing measures". Furthermore, in years 2020 and 2021, 155 International travel measures recorded. This is not a random fact, as most of the countries, including Greece, have banned air transportation high risk areas or taken special measures which were related to airports protocols [@sun2023; @damaskos2020].

```{r, message=FALSE, warning=FALSE, echo=FALSE}
 measure_category_by_year <- data %>%
                                group_by(`Entry year`) %>%
                                count(`Measure Category`)
  
  g1 <- ggplot(measure_category_by_year, aes(y=n, x=`Entry year` , fill=`Measure Category`))+
          geom_bar(stat = 'identity', position = 'dodge')+
          ggtitle("Category of Measures by Year") +
          theme(
                plot.title = element_text(hjust = 0.5)
          )+
          xlab("Year") +
          ylab("Number of Measures") +
          geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
          scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
          theme_hc()+
          scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  g1
```

Furthermore, the below graph shows the type of changes in the different categories of measures. Specially, it is observed that among all the categories the type of change: " A measure or type of measures is put in place for the first time" outperforms the other categories.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
changes <- data %>%
    dplyr::select(`Measure Category`,`Change in measure`)

  changes$`Change in measure`[changes$`Change in measure` == 0] <- 'Change of measure is not applicable' 
  changes$`Change in measure`[changes$`Change in measure` == 1] <- 'A measure or type of measure is put in place for the first time'
  changes$`Change in measure`[changes$`Change in measure` == 2] <- 'Measure is either re-introduced, extended or strengthened/intensified'
  changes$`Change in measure`[changes$`Change in measure` == 3] <- 'A measure is eased or lifted'
  
  changes_distribution <- changes %>%
    group_by(`Change in measure`) %>%
    count(`Change in measure`)
  
  g11 <- ggplot(changes, aes(fill=`Change in measure`, x = `Measure Category`))+
    geom_bar(stat='count', position = 'dodge')+
    ggtitle("Type of Change in each Measures' Category") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Measures Categories") +
    ylab("Count") +
    geom_text(stat = 'count', aes(label = ..count..), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")+
    theme(axis.text.x = element_text(angle = 18, vjust = 1, hjust = 1, size = 11))

  g11
```

Closing this section, it is presented a very useful interactive graph, in the case you want to explore all the categories and subcategories of anti-covid measures that applied in Greece.

```{r, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE}
measures_category_subcategory <- data %>%
    dplyr::select(`Measure Category`, `Measure Subcategory`, Measure)%>%
    group_by(`Measure Category`)
  
  
  g2_tree_plot_measures_categories <- collapsibleTree(measures_category_subcategory, c("Measure Category", "Measure Subcategory", "Measure"), root = "Category of Measures")
  g2_tree_plot_measures_categories
```

### Local Measures and Cities

Local Measures were necessary and very important to tackle the pandemic. As [@aspriadis2021][^1] supports, this level of enforcement can greatly assist a local community in limiting the spread of a virus. So, the below "wordcloud" graph show that in our capital city Athens, Thessaloniki and Mykonos implemented the majority of the measures.

[^1]: [@aspriadis2021] wrote based on the paper entitled: "Community crisis management lessons from Philadelphia's 1793 epidemic" from Edward J Glantz (2014).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
  local_measures_for_map <- read_xlsx("local_measures_for_map_4.xlsx")
  
  local_measures_for_map$`Region that measure adapted`[local_measures_for_map$`Region that measure adapted` == "Eordea"] <- 'Eordaea'
  
  # Add long and lat based on cities
  
  library(leaflet)
  library(leaflet.extras)
  library(tidygeocoder)
  
  # Wordcloud for cities
  library(ggwordcloud)
  
  set.seed(123)
  cities <- local_measures_for_map %>%
    dplyr::select(`Region that measure adapted`) %>%
    count(`Region that measure adapted`)
  
  ggwordcloud(cities$`Region that measure adapted`, cities$n)
```

Focusing on the local measures in the country, an interactive map created, in order to show the count of measures and the cities or more general territories which they applied depending on the level of enforcement.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9.5, fig.align='center'}
  local_measures_for_map_stats_final <- read_xlsx("local_measures_data_for_map.xlsx")
  
  leaflet(local_measures_for_map_stats_final) %>%
    addTiles() %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Measure is a requirement"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "darkred", library = "glyphicon"),
      group = "Measure is a requirement",
      popup = ~popup_info
    ) %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Not applicable to the measure"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "lightblue", library = "glyphicon"),
      group = "Not applicable to the measure",
      popup = ~popup_info
    ) %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Measure is a recommendation"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "darkgreen", library = "glyphicon"),
      group = "Measure is a recommendation",
      popup = ~popup_info
    ) %>%
    addControl(html = 'Count of Local Measures in Greece based on the Level of Enforcement', position = 'bottomleft') %>%
    addLayersControl(
      overlayGroups = c("Measure is a requirement", "Not applicable to the measure", "Measure is a recommendation"),
      options = layersControlOptions(collapsed = FALSE)
    )
```

### Scope of Measures

It is known that measures can be split into nationwide and local measures. Nationwide measures has more general character than local measures which aims to specific regions and can be differentiated from the first one [@conteduca2022]. In Greece, 528 measures classified as locals and 279 as nationwide. In addition, the majority both of local and nationwide measures were mandatory.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
data$`Scope of measure`[data$`Scope of measure` == 0] <- 'Local Measure'
  data$`Scope of measure`[data$`Scope of measure` == 1] <- 'Nationwide Measure'
  data$`Level of enforcement`[data$`Level of enforcement` == 0] <- 'Not applicable to the measure'
  data$`Level of enforcement`[data$`Level of enforcement` == 1] <- 'Measure is a recommendation'
  data$`Level of enforcement`[data$`Level of enforcement` == 2] <- 'Measure is a requirement'
  
  
  scope_of_measure <- data %>%
    dplyr::select(`Scope of measure`) %>%
    count(`Scope of measure`)
  
  g3 <- ggplot(scope_of_measure, aes(y=n, x=`Scope of measure` , fill=`Scope of measure`))+
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Scope of Measures") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Scope Category") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  
  scope_of_measure_by_level_of_enforcement <- data %>%
    dplyr::select(`Scope of measure`, `Level of enforcement`) %>%
    group_by(`Scope of measure`) %>%
    count(`Level of enforcement`)
  
  g3
  
  g4 <- ggplot(scope_of_measure_by_level_of_enforcement, aes(y=n, x=`Scope of measure` , fill=`Level of enforcement`))+
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Scope of Measures by Level of Enforcement") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Scope Category") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12") 
  
  g4
```

### Sources of Measures

Without any doubt, a wide variety of media, such as TV's news, information broadcast, websites, governmental websites and institutes and the Website of WHO have contributed significantly to the effort to deal with the pandemic Covid-19. The greek society informed the imposition of new measures or the changes of the existing by the mass media. In a parallel line, many information shows hosted scientists, such as doctors and epidemiologists, to better inform the the public about the pandemic and the measures against it [@anwar2020]. So, as it seems in the next graph, 475 measures' records based on Governmental websites, 209 on Media Articles and 123 on WHO's websites.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
data_for_links <- data %>%
    dplyr::select(`Entry ID`,Source, Links, `Date of access`)

  
  data_for_links$Source[data_for_links$Source == 1] <- 'Governmental Website'
  data_for_links$Source[data_for_links$Source == 2] <- 'Media Article'
  data_for_links$Source[data_for_links$Source == 3] <- 'WHO Country Office Communication'
  
  # Distribution of Measures' Records Sources
  
  g7 <- ggplot(data_for_links, aes(x= Source, fill = Source))+
    geom_bar(stat='count', position = 'dodge')+
    ggtitle("Sources of Measures' Records based on Source Type") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Source Type") +
    ylab("Count") +
    geom_text(stat = 'count', aes(label = ..count..), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  g7
```

Another interesting result is the fact that from governmental sources the most used websites were those of the [Greek Ministry of Health](https://www.moh.gov.gr/) (145 measures' records) and the [pandemic's official governmental website](https://covid19.gov.gr/) (131 measures' records). From the category of information websites, 141 measures' records based on [Kathimerini newspaper](https://www.ekathimerini.com/). These statistics are presented on the graph below.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.height=8}
 measures_sites <- read_xlsx('measures_sites.xlsx')

  g10 <- ggplot(measures_sites, aes(y=`Row Labels`, x=`Count of Sites`)) +
    geom_bar(stat='identity', position = 'dodge', fill = 'darkred')+
    ggtitle("Websites Distribution") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Count") +
    ylab("Websites") +
    geom_text(aes(label = `Count of Sites`), position = position_dodge(width = 0.9), hjust = -0.3, colour = "black") +
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  
  g10
```

### Conclusion

To sum up, it is well known that without the measures against Covid-19, we did not have face with the pandemic in an effective way. Statistics and facts that based on their data are the "compass" for a future possible pandemic, known as "Pandemic X" [@Iserson2020]. For sure, this hypothetical pandemic will make its own conditions and probably it will require finding new tactics to deal with it, but a look back to the history of Covid's-19 era, will be a good foothold.

### References
