---
title: "Measures Against Covid-19 in Greece: A Small Descriptive Analysis"
author: "Georgios Lamprou"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
bibliography: references.bib
---

```{r setup, include=FALSE}
# Set global options for all code chunks
knitr::opts_chunk$set(
  fig.width = 8,         # Default figure width in inches
  fig.height = 5.5,        # Default figure height in inches
  fig.align = "center"   # Center align all plots
)
```

```{=html}
<style>
  body {
    font-size: 16px; /* Increase the base font size */
    text-align: justify;
  }
</style>
```

### Introduction

Five years after the start of the Covid-19 pandemic, the world continues to live in a situation of symbiosis with the virus Sars-Cov-2. Even in our days, a lot of people suffer from coronavirus for first or second or third time, a phenomenon which called reinfection [@reinfect2024]. However, in the difficult triennial (2020 - 2023), governments around the world were constantly making decisions to introduce or revise a wide variety of measures in order to mitigate the pandemic and protect the human life [@güner2021].

The same approach was adopted in Greece with general effectiveness. Other nations admired the Greek effort and tactics to deal with the pandemic. In any circumstance, the immediate strict measures' application on a nationwide as well as on a local level and the discipline of people played a huge role for this kind of success [@papathanassopoulos2023].

In this context, this article aims to present a small descriptive analysis of the measures against Covid-19 in Greece. The analysis is divided into three sections:

1.  Measures During the Pandemic Years

2.  Categorization of Anti-Covid-19 Measures

3.  Local and Nationwide Measures

### Tools

The visualizations created with R Programming Language and the technology of RMarkdown Documents used to write the article. To view the full code in R of this work, follow the next link: [https://bitly.cx/N4MR](https://www.linkedin.com/redir/suspicious-page?url=https%3A%2F%2Fbitly%2ecx%2FN4MR). Furthermore, the database "Public Health and Social Measures in Response to COVID-19" from World Health Organization (WHO) is used for the acquisition of data. To explore this useful database for European Region, the corresponding link is provided here: [https://phsm.euro.who.int/covid-19](#0).

### Measures During the Pandemic Years

To begin with, as seems in the graph below, the majority of the measures recorded on 2021 (393 records) and the minority of them on 2024 (3 records).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Libraries
  library(dplyr)
  library(readxl)
  library(ggplot2)
  library(ggthemes)
  library(plotly)
  library(collapsibleTree)
  library(paletteer)
  library(lisa)
  library(ggrepel)
  library(patchwork)
  library(GGally)

  data <- read_xlsx("phsm_database_04032024_greece.xlsx", col_names = TRUE)
  
  # Data Processing and Transformations
  
  data$`Start of measure` <- as.Date.POSIXct(data$`Start of measure`)
  data$`End of measure` <- as.Date.POSIXct(data$`End of measure`)
  data$`Date of access` <- as.Date.POSIXct(data$`Date of access`)
  
  
  # Different Statistics - Graphs
  
  # Number of Measures' Record during the years
  
  freq_of_measures_based_on_year_entry <- data %>%
    dplyr::select(`Entry year`)%>%
    count(`Entry year`)
  
  freq_of_measures_based_on_year_entry$`Entry year` <- as.factor(freq_of_measures_based_on_year_entry$`Entry year`)
  
  g0 <- ggplot(freq_of_measures_based_on_year_entry, aes(x= `Entry year`, y=n, fill=`Entry year`)) +
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Records of Measures by Year") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Year") +
    ylab("Number of Records") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
g0  
```

This point is not strange. Instead, is very logical for 2 main reasons:

1.  2021 was the next year of 2020, the pandemic's official starting year [@jebril2020] while an important amount of 2021's days included in the second lockdown which started in November 2020 and finished 210 later [@papazisis2021].

2.  In 2021 many measures which were related to vaccinations and Delta mutation of Sars-Cov-2 virus, were introduced for 1st time or modified, [@Liossi2023; @Tsiambas2021].

The above result could also be framed by the following time series graph. In this, it is observed that the most of the measures started in 2021.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
  data_for_timeline_of_measure_started <- data %>%
    dplyr::select(`Start of measure`) %>%
    count(`Start of measure`)
  
  data_for_timeline_of_measure_ended <- data %>%
    dplyr::select(`End of measure`) %>%
    count(`End of measure`)
  
  g5 <- ggplot(data_for_timeline_of_measure_started, aes(x=`Start of measure`, y=n))+
    geom_line(stat = 'identity',color='darkred') +
    ggtitle("Total Number of Measures Started by Day") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Date") +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12") 
   
  ggplotly(g5)
```

### Categorization of Preventive Measures

In the effort to place all the preventive measures against Covis-19 into categories, the next bar plot, shows that in the triennial 2020, 2021, 2022 most of the measures belong to the category "Social and physical distancing measures". Furthermore, in years 2020 and 2021, 155 International travel measures recorded. This is not a random fact, as most of the countries, including Greece, have banned air transportation high risk areas or taken special measures which were related to airports protocols [@sun2023; @damaskos2020].

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11}
 measure_category_by_year <- data %>%
                                group_by(`Entry year`) %>%
                                count(`Measure Category`)
  
  g1 <- ggplot(measure_category_by_year, aes(y=n, x=`Entry year` , fill=`Measure Category`))+
          geom_bar(stat = 'identity', position = 'dodge')+
          ggtitle("Category of Measures by Year") +
          theme(
                plot.title = element_text(hjust = 0.5)
          )+
          xlab("Year") +
          ylab("Number of Measures") +
          #geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
          scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
          theme_hc()+
          scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  ggplotly(g1)
```

Furthermore, the below graph shows the distribution of anti-covid measures that put in place for the first time. So, it seems that 203 measures belong to the category: "Social and physical distancing measures", 99 in "International travel measures" and 27 measures fall under "Surveillance and response measures" and "Biological Measures" equally.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=13}
changes <- data %>%
    dplyr::select(`Measure Category`,`Change in measure`)

  changes$`Change in measure`[changes$`Change in measure` == 0] <- 'Change of measure is not applicable' 
  changes$`Change in measure`[changes$`Change in measure` == 1] <- 'A measure or type of measure is put in place for the first time'
  changes$`Change in measure`[changes$`Change in measure` == 2] <- 'Measure is either re-introduced, extended or strengthened/intensified'
  changes$`Change in measure`[changes$`Change in measure` == 3] <- 'A measure is eased or lifted'
  
  first_measures_changes_distribution <- changes %>%
    group_by(`Measure Category`) %>%
    dplyr::filter(`Change in measure` =='A measure or type of measure is put in place for the first time')%>%
    count(`Change in measure`)
  
  g11 <- ggplot(first_measures_changes_distribution, aes(y= `Measure Category`, x = n, fill = `Measure Category`))+
    geom_bar(stat='identity', position = 'dodge')+
    ggtitle("Measures that Intorduced for First Time by Measure Category") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Count") +
    ylab("Measure Category") +
    #geom_text(aes(label = n), position = position_dodge(width = 0.9), hjust = -0.3, colour = "black", size = 5)+
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")

  ggplotly(g11)
```

Focusing on this category of measures, the below graph answers the question of what were the changes to the measures applied for the first time. So, it is observed that in all the categories the majority of measures have been changed in terms of re-introduction, extension or intensity.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=13}
first_measures_changes <- changes %>%
    group_by(`Measure Category`) %>%
    dplyr::filter(`Change in measure` !='A measure or type of measure is put in place for the first time')
  
g12 <- ggplot(first_measures_changes, aes(y= `Measure Category`, fill = `Change in measure`))+
    geom_bar(stat='count', position = 'dodge')+
    ggtitle("Distribution of Measures that Introduced for the First Time based on Type of Changes") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )+
    xlab("Count") +
    ylab("Measure Category") +
    #geom_text(stat = 'count', aes(label = ..count..), position = position_dodge(width = 0.9), hjust = -0.3, colour = "black", size = 5)+
    theme_hc()+
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")

ggplotly(g12)
```

Closing this section, it is presented a very useful interactive graph, in the case you want to explore all the categories and subcategories of anti-covid measures that applied in Greece.

```{r, fig.height=8, message=FALSE, warning=FALSE, echo=FALSE}
measures_category_subcategory <- data %>%
    dplyr::select(`Measure Category`, `Measure Subcategory`, Measure)%>%
    group_by(`Measure Category`)
  
  
  g2_tree_plot_measures_categories <- collapsibleTree(measures_category_subcategory, c("Measure Category", "Measure Subcategory", "Measure"), root = "Category of Measures")
  g2_tree_plot_measures_categories
```

### Local and Nationwide Measures

It is known that measures can be split into nationwide and local measures. From the one hand, nationwide measures has more general character than local measures which aims to specific regions and can be differentiated from the first one [@conteduca2022]. From the other hand, local Measures were necessary and very important to tackle the pandemic. As [@aspriadis2021][^1] supports, this level of enforcement can greatly assist a local community in limiting the spread of a virus. So, the below "wordcloud" graph show that in our capital city Athens, Thessaloniki and Mykonos implemented the majority of the measures.

[^1]: [@aspriadis2021] wrote based on the paper entitled: "Community crisis management lessons from Philadelphia's 1793 epidemic" from Edward J Glantz (2014).

```{r, message=FALSE, warning=FALSE, echo=FALSE}
  local_measures_for_map <- read_xlsx("local_measures_for_map_4.xlsx")
  
  local_measures_for_map$`Region that measure adapted`[local_measures_for_map$`Region that measure adapted` == "Eordea"] <- 'Eordaea'
  
  # Add long and lat based on cities
  
  library(leaflet)
  library(leaflet.extras)
  library(tidygeocoder)
  
  # Wordcloud for cities
  library(ggwordcloud)
  
  set.seed(123)
  cities <- local_measures_for_map %>%
    dplyr::select(`Region that measure adapted`) %>%
    count(`Region that measure adapted`)
  
  ggwordcloud(cities$`Region that measure adapted`, cities$n)
```

Focusing on the local measures in the country, an interactive map created, in order to show the count of measures and the cities or more general territories which they applied depending on the level of enforcement.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=9.5, fig.align='center'}
  local_measures_for_map_stats_final <- read_xlsx("local_measures_data_for_map.xlsx")
  
  leaflet(local_measures_for_map_stats_final) %>%
    addTiles() %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Measure is a requirement"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "darkred", library = "glyphicon"),
      group = "Measure is a requirement",
      popup = ~popup_info
    ) %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Not applicable to the measure"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "lightblue", library = "glyphicon"),
      group = "Not applicable to the measure",
      popup = ~popup_info
    ) %>%
    addAwesomeMarkers(
      data = subset(local_measures_for_map_stats_final, `Level of enforcement` == "Measure is a recommendation"),
      ~longitude, ~latitude,
      icon = awesomeIcons(icon = "map-marker", markerColor = "darkgreen", library = "glyphicon"),
      group = "Measure is a recommendation",
      popup = ~popup_info
    ) %>%
    addControl(html = 'Count of Local Measures in Greece based on the Level of Enforcement', position = 'bottomleft') %>%
    addLayersControl(
      overlayGroups = c("Measure is a requirement", "Not applicable to the measure", "Measure is a recommendation"),
      options = layersControlOptions(collapsed = FALSE)
    )
```

To continue this section in a more general context, the next graph shows that in Greece, 528 measures classified as locals and 279 as nationwide. In addition, the majority of both of local and nationwide measures were mandatory.

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=11}
data$`Scope of measure`[data$`Scope of measure` == 0] <- 'Local Measure'
  data$`Scope of measure`[data$`Scope of measure` == 1] <- 'Nationwide Measure'
  data$`Level of enforcement`[data$`Level of enforcement` == 0] <- 'Not applicable to the measure'
  data$`Level of enforcement`[data$`Level of enforcement` == 1] <- 'Measure is a recommendation'
  data$`Level of enforcement`[data$`Level of enforcement` == 2] <- 'Measure is a requirement'
  
  
  scope_of_measure <- data %>%
    dplyr::select(`Scope of measure`) %>%
    count(`Scope of measure`)
  
  g3 <- ggplot(scope_of_measure, aes(y=n, x=`Scope of measure` , fill=`Scope of measure`))+
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Scope of Measures") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Scope Category") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12")
  
  scope_of_measure_by_level_of_enforcement <- data %>%
    dplyr::select(`Scope of measure`, `Level of enforcement`) %>%
    group_by(`Scope of measure`) %>%
    count(`Level of enforcement`)
  
  g3
  
  g4 <- ggplot(scope_of_measure_by_level_of_enforcement, aes(y=n, x=`Scope of measure` , fill=`Level of enforcement`))+
    geom_bar(stat = 'identity', position = 'dodge')+
    ggtitle("Scope of Measures by Level of Enforcement") +
    theme(
                plot.title = element_text(hjust = 0.5)
          )+
    xlab("Scope Category") +
    geom_text(aes(label = n), position = position_dodge(width = 0.9), vjust = -0.3, colour = "black") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_hc() +
    scale_fill_paletteer_d("ggthemes::Classic_Blue_Red_12") 
  
  g4
```

### Conclusion

To sum up, it is well known that without the measures against Covid-19, we did not have face with the pandemic in an effective way. Statistics and facts that based on their data are the "compass" for a future possible pandemic, known as "Pandemic X" [@Iserson2020]. For sure, this hypothetical pandemic will make its own conditions and probably it will require finding new tactics to deal with it, but a look back to the history of Covid's-19 era, will be a good foothold.

### References
